### 2014-10-02 paper/include/background.tex
made some placeholders for new subsections

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.

Trends - Storage array rectangular
            cant go wider (word lines length = timing = power)
            gets taller = more refreshes
            over time .... projektions (see Intel)

>>>>

\todo[inline]{Drafty stuff below}

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access. MENTION TIMING ISSUES IF ROWS GET LONGER.

When the DRAM chips become larger the throughput loss increases exponentially as \reffig{fig:dram_throughput} shows. Also when additional rows are added to the chip, more refresh commands has to be sent and of course gives this many more actual refreshes, especially if CBR is used. More refreshes means larger power consumption and is is also so that the refresh bocomes a growing portion of the total DRAM energy consumption. In \reffig{fig:dram_power} is is seen that when the device capacity reaches 64 GB the refresh power will nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

In the next years it is projected that ... blablabla [source] so it is important and will be more important in the future to lower the total DRAM power consumption and an relatively easy way to do this is to optimizing the DRAM refresh. Next. we present a bunch of techniques that target DRAM refresh optimization.

### 2014-10-02 paper/include/reducing_refresh_power.tex
cleaned
added retention techs to sec 3
more reorganization
made some placeholders for new subsections

\section{Reducing refresh power} 
\label{sec:red}
Roadmap 
Worth mentioning scheduling real acces togheter with refresh ops
Write an intro to the ARTV approches.
There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon. The four sort of information are: access recency, cell retention times, error tolerance of the data, and line validity. This categorization follows the taxonamy proposed by authors \cite{dtail}.

\subsection{Types of refreshing}

@@

\subsubsection*{\textbf{Retention Time}}
Retention time refers to the period during which cells hold valid values as charge gradually leaks. Process variation [10][16] causes the retention time of DRAM cells to vary across the chip. Note that approaches that ex-
ploit R information are insensitive to system workloads and global memory usage, which makes them attractive ompo
nents for a refresh-optimized memory subsystem.

@@

\subsection{Reducing refresh power approches}

\subsection{Access Recency}

>>>>

\section{Reducing refresh power} 
\label{sec:red}
\todo[inline]{Do an intro that includes a roadmap, mention the refresh scheduling approach that schedules real acces togheter with refresh ops. Make an intro to ARTV (somewhat done) }
There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon. The four sort of information are: access recency, cell retention times, error tolerance of the data, and line validity. This categorization follows the taxonamy proposed by authors \cite{dtail}.

\subsection{Approaches to reduce refresh power}
\todo[inline]{Replace the copy-paste texts}

@@

\subsubsection*{\textbf{Retention Time}}
Retention time refers to the period during which cells hold valid values as charge gradually leaks. Process variation [10][16] causes the retention time of DRAM cells to vary across the chip. Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.

@@

\subsection{Techniquies to reducing refresh power}

\subsubsection*{\textbf{Smart Refresh}}
\label{par:smartrefresh}
\input{include/smart_refresh}

\subsubsection*{\textbf{Refrint}}
\label{par:refrint}
\input{include/refrint}

\subsubsection*{\textbf{DTail-RV}}
\label{par:dtail}
\input{include/dtail}

\subsubsection*{\textbf{RAIDR}}
\label{par:raidr}
\input{include/raidr}

\subsubsection*{\textbf{RIO and PARIS}}
\label{par:rioparis}
\input{include/rioparis}

\subsubsection*{\textbf{SECRET}}
\label{par:secret}
\input{include/secret}

\subsubsection*{\textbf{ESKIMO}}
\label{par:eskimo}
\input{include/eskimo}

\subsubsection*{\textbf{Sparkk}}
\label{par:sparkk}
\input{include/sparkk}

\subsubsection*{\textbf{Flikker}}
\label{par:flikker}
\input{include/flikker}

\subsubsection*{\textbf{CREAM}}
\label{par:cream}
\input{include/cream}

\subsubsection*{\textbf{Energy Optimization in 3D MPSoCs with Wide-I/O DRAM
Using Temperature Variation Aware Bank-wise Refresh}}
\label{par:tempaware}
\input{include/tempaware}

\subsection{Differences and similarities of the approaches}

Differences and similarities for ARTV approaches.

What in the DRAM devices is modified? / Where is data placed.

How much knowledge does the techniques have about the running software?

### 2014-10-02 paper/include/smart_refresh.tex
made some placeholders for new subsections

The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.

>>>>

\todo[inline]{Make the text flow}
The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.

### 2014-10-02 paper/include/sparkk.tex
made some placeholders for new subsections

\section{Tolerance} 
\label{sec:tol}

Papers that use this technique:

Flikker\cite{flikker}

Sparkk\cite{sparkk}
Multiple authors have suggested techniques where programmers can specify which data is critical and can not tolerate any bit errors and which data can be stored approximately. However, in these approaches all bits in the approximate area are treated as equally important. We show that this produces suboptimal results and higher energy savings or better quality can be achieved, if a more fine-grained approach is used. Our proposal is able to save more refresh power and enables a more effective storage of non-critical data by utilizing a non-uniform refresh of multiple DRAM chips and a permutation of the bits to the DRAM chips.

Refrint\cite{refrint}

>>>>

%Sparkk
Tolerance!

\todo[inline]{Ccopy-paste:}
Multiple authors have suggested techniques where programmers can specify which data is critical and can not tolerate any bit errors and which data can be stored approximately. However, in these approaches all bits in the approximate area are treated as equally important. We show that this produces suboptimal results and higher energy savings or better quality can be achieved, if a more fine-grained approach is used. Our proposal is able to save more refresh power and enables a more effective storage of non-critical data by utilizing a non-uniform refresh of multiple DRAM chips and a permutation of the bits to the DRAM chips.
